Temporal Continuity and the Judgment of Actual Causation

### 2021-01-22 docs/resources/CogSci_paper/paper.tex
added to discussion

%\title{A continuity-based account of actual causation}
%\title{Identifying actual causes by tracking Changes of State in Continuous Time}
\title{Actual Causes as Changes of State in Continuous Time}

@@

\textbf{Keywords:} 
\end{abstract}

@@

Outline:

\textbf{I}. Summary and implications of the results

\textbf{II}. To say why these results cannot really be explained by the current theories of causation and why our theory can better account for them. I don't know if I should talk more about CF and PP accounts than I do in introduction, like saying something like that is probably not useful:

[\textit{According to the CF account, the general idea is that an event A is said to be a cause of a distinct event B if the occurrence of A makes a difference in the occurrence of B. More precisely A is a cause of B if and only if A and B are true, and A hadn't occurred B wouldn't have occurred. This interpretation calls upon counterfactual scenario or possible worlds where the presumed cause of the effect is removed from the system while all other relevant factors are kept as unchanged as possible compared to the actual world. According to us the main problem of this interpretation is that its best models explicitely rejects the idea of grounding causation on temporal information.
According to the PP account, an event A causes B when there is a physical connection between them. In one of its latest and probably most convincing formulation, this theory distinguishes two things : causal process and causal interaction (i.e. causation). A causal process is a physical process involving an object which conserves a certain quantity, like mass-energy, accross space and time. A causal interaction or causation is an exchange of that conserved quantity. However the theory goes sometimes against some of our best causal intuitions and to makes implicitely use of the CF analysis to explain our judgements: if I hold the head of my ennemy under water and make him die, I'm not genuinely the cause of his death; rather I'm actually preventing the possibility of a genuine causation which is breathing oxygen in order to live.}]

Or if I do have to explicit both theories I should insist on what the predictions of both theories are for our experiments and why they cannot explain people's intuition (the role of time).

\textbf{III}. To insist, maybe, on the work of Glymour where our project in part originated: explaining his caveat about treating events as pure propositions. To explain how our experiments extend this idea and insist more on the role of changes in continuous time. Maybe we should present some insights about how we could formalize further the model? How we could use formal tools like Graphical models?

\textbf{IV}. To present further experiments that would be interesting to make (with loops, cases of prevention, etc.)

>>>>

%\title{A continuity-based account of actual causation}
%\title{Identifying actual causes by tracking Changes of State in Continuous Time}
%\title{Actual Causes as Changes of State in Continuous Time}
\title{Temporal Continuity and the Judgment of Actual Causation}

@@

abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract 
abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract abstract

\textbf{Keywords: causal inference; causal selection; causal explanation} 
\end{abstract}

@@

%Outline:
%
%\textbf{I}. Summary and implications of the results
%
%\textbf{II}. To say why these results cannot really be explained by the current theories of causation and why our theory can better account for them. I don't know if I should talk more about CF and PP accounts than I do in introduction, like saying something like that is probably not useful:
%
%[\textit{According to the CF account, the general idea is that an event A is said to be a cause of a distinct event B if the occurrence of A makes a difference in the occurrence of B. More precisely A is a cause of B if and only if A and B are true, and A hadn't occurred B wouldn't have occurred. This interpretation calls upon counterfactual scenario or possible worlds where the presumed cause of the effect is removed from the system while all other relevant factors are kept as unchanged as possible compared to the actual world. According to us the main problem of this interpretation is that its best models explicitely rejects the idea of grounding causation on temporal information.
%According to the PP account, an event A causes B when there is a physical connection between them. In one of its latest and probably most convincing formulation, this theory distinguishes two things : causal process and causal interaction (i.e. causation). A causal process is a physical process involving an object which conserves a certain quantity, like mass-energy, accross space and time. A causal interaction or causation is an exchange of that conserved quantity. However the theory goes sometimes against some of our best causal intuitions and to makes implicitely use of the CF analysis to explain our judgements: if I hold the head of my ennemy under water and make him die, I'm not genuinely the cause of his death; rather I'm actually preventing the possibility of a genuine causation which is breathing oxygen in order to live.}]
%
%Or if I do have to explicit both theories I should insist on what the predictions of both theories are for our experiments and why they cannot explain people's intuition (the role of time).
%
%\textbf{III}. To insist, maybe, on the work of Glymour where our project in part originated: explaining his caveat about treating events as pure propositions. To explain how our experiments extend this idea and insist more on the role of changes in continuous time. Maybe we should present some insights about how we could formalize further the model? How we could use formal tools like Graphical models?
%
%\textbf{IV}. To present further experiments that would be interesting to make (with loops, cases of prevention, etc.)

We presented an account of actual causation that highlights the role of temporal continuity and described experiments that support two of its predictions. First, people tend to identify root causes rather than immediate causes as the main cause of an effect. Second, when an effect is produced by the convergence of multiple causal pathways, participants tend to identify the pathway that has temporal continuity with the effect as the primary cause.

Because the continuity account considers changes of state in continuous time, it can exploit a ``no coincidences'' principle to identify a single main cause of an effect. This principle applies broadly to causation in physical systems, but is less applicable to social scenarios such as voting scenarios~\cite{livengood13} Exceptions to the ``no coincidences'' principle are possible even for physical systems, and the continuity account can explain the illusory perception of actual causation that occurs in some such cases~\cite{thorstadw16}. For example, an 11 year old boy whacked a stick against a telephone pole at the instant at which the Great Northeastern Blackout hit his town, and ran home terrified that he had caused the power outage~\cite{gelbr1965night}.

The continuity account highlights ideas such as the ``no coincidences'' principle that go beyond default formulations of the counterfactual and physical process accounts of causation. Key predictions of the account, however, can be reconstructed within both of these general frameworks. The counterfactual account can exploit fine-grained temporal information if each event is supplemented with a timestamp. If the effect to be explained in Figure~\ref{fig:intro} ``the activation of the GCM at exactly 10 sec after the start of the trial,'' then the counterfactual account suggests that \ev{R}{C} is a cause of the effect but \ev{R}{D} is not, because if \ev{R}{D} had been different (e.g.\ by occurring slightly earlier or later) then the effect would still have occurred at the same instant.

As suggested earlier, standard formulations of the physical process view treat both \ev{R}{C} and \ev{R}{D} as causes of the effect in Figure~\ref{fig:intro}. Our data are consistent with the possibility that people view \ev{R}{C} and \ev{R}{D} as equally valid causes, but pick \ev{R}{D} when forced to break the tie for reasons that may be relatively superficial. For example, perhaps \ev{R}{D} is preferred because the activation on the continuous branch is continuous both in space and time, and therefore more prototypical of a causal process than the activation along the delay branch. We have planned a future study that asks whether the \ev{R}{C} over \ev{R}{D} is fundamental or superficial using a paradigm in which participants rank multiple candidate causes in order of importance, and in which ties are permitted.

% Other future studies
